---
title: "Comparing US Wheat Yield with Middle-Aged Marriages in Salzburg"
author: "Helmuth Breitenfellner"
date: "19.04.2019"
output:
  pdf_document:
classoption: "10pt"
subtitle: "Data Experiment for Data Stewardship"
header-includes:
- \usepackage{fancyhdr}
- \usepackage{lastpage}
- \usepackage{multicol}
- \pagestyle{fancy}
- \lhead{Helmuth Breitenfellner}
- \chead{}
- \lhead{https://orcid.org/0000-0003-4523-0945}
- \rhead{Helmuth Breitenfellner}
- \rfoot{Page \thepage{} of \pageref{LastPage}}
- \renewcommand{\footrulewidth}{0.4pt}
- \lfoot{Data Stewardship}
- \cfoot{}
- \newcommand{\twocolstart}{\begin{multicols}{2}}
- \newcommand{\twocolend}{\end{multicols}}
- \newcommand{\abstractstart}{\begin{abstract}}
- \newcommand{\abstractend}{\end{abstract}}
bibliography: references.bib
---
\abstractstart
This paper analyses the correlation between yield
of wheat in the United States on one side,
and the number of marriages by middle-aged (35-44)
men and women in Salzburg.

The source of the project can be found on GitHub:
https://github.com/helmuthb/dmp-exercise1.
All software and the report is licensed under MIT license.
The data is licensed as by the data providers.

* CC0: US Wheat Production Timetable [@wheat2019]
* CC BY 4.0: Age of partners at marriage [@salzburg2019]

This data experiment is the result of exercise 1
of the lecture "Data Stewardship".
\vspace{5mm}

\abstractend
\twocolstart

# Data Gathering

The data for [@salzburg2019] is available for
download from the Austrian
Data portal (https://www.data.gv.at),
and the data for [@wheat2019] from the US
Department of Agriculture, Economic Research
Service (https://www.ers.usda.gov/).

They have been downloaded and are available as part
of the repository in the folder `data/source`.

# Data Processing

The data from Salzburg is available in CSV or JSON
format. For the purpose of this experiment the CSV
version has been taken.

The data from the USDA is available as an Excel sheet.

For processing of the data the language $R$ is used,
which allows both reading of various data sources,
statistical analysis (like cross correlation), and
plotting of data for visual analysis and comparison.

The data files are read in, using `read.csv2` for
[@salzburg2019] and `read_excel` for [@wheat2019].

For [@salzburg2019] the age classes are transformed
into numerical values (the lowest age in each class),
and the two columns `wives` and `husbands` are
re-arranged as two lines with one column `spouses`
and `gender` ("female" or "male").
The age range is then filter for the range of
interest (35-44 years) and added up.

For [@wheat2019] only a range of the second sheet
is read (the region corresponding to wheat).
The field `year` is adjusted to numerical values.
Originally it contained always two consecutive
years, like 1972/1973, to indicate the harvest
season.
After the adjustment it contains only the second year.

The two data sets were then merged by `year` and stored
as a CSV file in `data/processed/us-wheat-sbg-marriages.csv`.

```{r,include=FALSE}
library(readxl)

# read data from Salzburg
salzburg <- read.csv2("data/source/marriage-szg-sex-age.csv",
                      stringsAsFactors = TRUE,
                      colClasses =
    c("NULL", "NULL", "factor", rep("numeric", 3) ),
                      col.names =
    c("NULL", "NULL", "age", "year", "wives", "husbands"),
                      skip = 2)


# adjust age levels: instead of long strings take
# youngest age of class
levels(salzburg$age) <- c(15, 20, 25, 30, 35,
                          40, 45, 50, 55, 60,
                          65, 70)
# and convert to integer
salzburg$age <- as.numeric(as.character(salzburg$age))

# sum up middle-aged (35-44) wives & husbands
library(dplyr)

salzburg <- salzburg %>%
  filter(age >= 35 & age <= 44) %>%
  group_by(year) %>%
  summarize(wives = sum(wives),
            husbands = sum(husbands))

# Reshape: one column "spouses", with
# a flag "gender" for husband or wife
wives <- data.frame(year = salzburg$year,
                    spouses = salzburg$wives,
                    gender = "female")
husbands <- data.frame(year = salzburg$year,
                       spouses = salzburg$husbands,
                       gender = "male")

salzburg <- rbind(wives, husbands)

# read data from USDA
wheat <- read_excel("data/source/Wheat_Data-All_Years.xls",
                    sheet = 2,
                    range = "B5:G157",
                    col_names =
    c("year", "a_p", "a_h", "production", "yield", "p"),
                    na = "--")
# adjust year column to numeric
# since the source has yyyy/yyyy+1 as season, the
# value is taken as yyyy+1
wheat$year = as.numeric(substr(wheat$year, 1, 4)) + 1

# merge data by year
combined <- merge(salzburg, wheat, by = "year")

# write data file back as CSV
write.csv(combined, "output/us-wheat-sbg-marriages.csv")

# split male and female for easier analysis
male <- combined %>%
  filter(gender == "male")
female <- combined %>%
  filter(gender == "female")
```

# Data Visualization

```{r, include=FALSE}
options(digits = 3)
```
```{r,echo=FALSE}
# Plot spouses vs yield
library(ggplot2)
combined %>%
  ggplot(aes(x = yield, y = spouses, color = gender)) +
  geom_point() +
  geom_smooth(method = "lm")
```

The previous plot shows wheat yield in comparison to
weddings in Salzburg.
The corresponding linear regression line has been
added for enhanced visualization.

More specifically, looking at the correlation
coefficient between husbands in the middle age range
(35 - 44 years) and wheat yield gives a coefficient
of `r cor(female$spouses, female$yield)` for women
and `r cor(male$spouses, male$yield)` for men.

This indicates a high level of correlation.

# Technical Description of Experiment

The following $R$ packages have been used in this
data experiment:

* `readxl` - for reading Excel sheet
* `dplyr` - for data manipulation
* `ggplot2` - for graphical plots
* `DiagrammeR` - for UML diagrams
* `DiagrammeRsvg` - for UML diagrams
* `rsvg` - for including UML diagrams

The following chart shows the steps of the experiment:

```{r,echo=FALSE}
library(DiagrammeR)
library(DiagrammeRsvg)
library(rsvg)
library(knitr)
uml <- DiagrammeR::grViz("
digraph {
	  # Nodes
	  node [shape = diamond]
    salzburg [shape = rectangle, label = 'Marriages Salzburg']
    wheat[shape = rectangle, label = 'Wheat US']
    salzburg_processing[label = 'Data Reshape']
    wheat_processing[label = 'Range Selection']
    merging[label = 'Data Merging']
    merged[shape = rectangle, label = 'Merged Dataset']
    packages[label = 'R Packages']
    plot[shape = rectangle, label = 'Correlation Plot']

	  # Edges
	  edge [color = black,
	        arrowhead = vee]
	  rankdir = TD
	  wheat->wheat_processing [label = 'Input']
	  salzburg->salzburg_processing [label = 'Input']
	  wheat_processing->merging
	  salzburg_processing->merging
    merging->merged
    merged->packages
    packages->plot

	{rank=same; wheat; salzburg}
	{rank=same; wheat_processing; salzburg_processing}
	{rank=same; merging; merged}
  {rank=same; packages; plot}
	}
")
tmp <- capture.output(rsvg_pdf(charToRaw(export_svg(uml)), 'uml.pdf'))
include_graphics("uml.pdf")
```

A `Dockerfile` is provided together with the required $R$ packages
and the
\LaTeX
libraries used for creating this report.
The Docker image is created on Docker Hub and can be used pre-compiled.

# Further Analysis

Clearly one can hardly explain this correlation
between marriages in Salzburg and wheat yield
in the United States.
So what is the reason for this correlation then?

When looking through the data set more in detail,
it becomes obvious that the actual correlation
is in the year.
Both the yield of wheat in the United States
is increasing, as is the number of marriages of
middle-aged persons in Salzburg.

These correlations are even stronger, the corresponding
coefficients are
`r cor(female$spouses, female$year)` for women,
`r cor(male$spouses, male$year)` for men, and
`r cor(male$yield, male$year)` for wheat yield.

```{r,echo=FALSE}
# Plot yield vs year
library(ggplot2)
combined %>%
  ggplot(aes(x = year, y = yield)) +
  geom_point() +
  geom_smooth(method = "lm")
```

```{r,echo=FALSE}
# Plot spouses vs year
library(ggplot2)
combined %>%
  ggplot(aes(x = year, y = spouses, color = gender)) +
  geom_point() +
  geom_smooth(method = "lm")
```

\twocolend

# References
